Comprehensive Algorithmic Resilience for Scientific Applications

### 2013-01-16 paper.tex
Some more text

\title{Comprehensive Algorithmic Resilience for Scientific Applications}
\author{Sui Chen, 
%\affil{Louisiana State University}
Greg Bronevetsky, Marc Casas-Guix
%\affil{Lawrence Livermore National Laboratory}
Lu Peng,
%\affil{Louisiana State University}
}

@@

\subsection{Alternating Direction Method of Multipliers (ADDR)}
\label{sec:apps:lasso}
The ADDR method solves under-unconstrained linear problems $Ax=b$ for $x$ ($A$ has fewer rows than columns) while minimizing the function $f(x)$.
This is an iterative algorithm that spends most of its time in the linear algebra operations, matrix-matrix multiplication, matrix-vector multiplication, rank-k update and Cholesky factorization, using the implementations of these routines from the GNU Scientific Library~\cite{gsl:2011}.
We focused on a specific application that uses the ADDR method to solve Lasso problems, which are defined as minimizing the function $\frac{1}{2} \left|| Ax - b \right||_2^2 + \lambda*\left|| x \right||_1$.
In our experiments we considered linear systems of size \textbf{???} that were generated \textbf{???}.

@@

The DRC program generates filters for HiFi audio systems to compensate for the reflection of sounds in a room, using impulse response measurements of the audio equipment and considering the positions of the listeners.
Although the algorithm is divided into many phases the most computationally intensive step consists of the GSL's implementation of the Fast Fourier Transform.
\greg{What inputs?}

@@

\paragraph{Matrix-vector multiplication (MVM)}:
The MVM algorithm $Ax=b$ is checked using a similar identity $(x^TA)x = x^T(Ax) = x^Tb$.
The product $x^TA$ is a checksum of the columns of matrix $A$ and when $x$ is the vector of 1's, the value used in our experiments, it is the sum of $A$'s columns.
Although the computation of $x^TA$ involves additions rather than multiplications the complexity of computing $x^TA$ is $O(n^2)$, the same as the original MVM operation, making this check significantly less efficient.
While this cost can be reduced significantly by computing $x^TA$ once and reusing it in subsequent multiplications by the same matrix, we did not consider it in this paper because MVM took up little application execution time.

@@

\paragraph{Fast Fourier Transform (FFT)}:
The FFT algorithm decomposes a function into a sum of sine waves of different frequencies: $f(x) = \sum_{n=0}^{N-1} x_n e^{-i2\pi k n / N}$ for some constant $k$.
We check the results of FFT by using Parseval's theorem, which states that $\sum_{n=0}^{N-1} \left| x[n] \right|^2 = \frac{1}{N} \sum_{k=0}^{N-1} \left| X[k] \right|^2$, where $x$ is the original function and $X$ is its transform.
This means that the total energy of the original function is preserved by the transform and provides an $O(n)$ time check for the $O(n log(n))$ FFT algorithm.

\paragraph{Runge-Kutta PDE Solver (RK)}:
This algorithm uses the \sui{4-th order} Runge-Kutta method for solving Ordinary Differential Equations of the form $\frac{dy}{dx} = f(y, x)$.
It works by advancing the variable $x$ by steps of size $h$ and using estimates of the derivative $\frac{dy}{dx}$ at each point $x$ to find the value of $y$ at the next point $x+h$.
The RK method in GSL uses adaptive step-size control where it simultaneously performs the RK method with two step sizes $h$ and $\frac{h}{2}$ (more precise).
If the difference between the results obtained with the two step sizes are larger than a threshold, the algorithm switches to the smaller step size to maintain accuracy and resumes.
Since the adaptive step control algorithm can also detect errors we utilize it to this end in our work.

@@

We now evaluate the effectiveness of our algorithmic detectors when errors are injected.
Figure~\ref{fig:algo_err_dist} shows the fraction of entries in the outputs of each routine (y-axis) that have an error of a given magnitude (x-axis) as error magnitudes span from ??? to ??? and errors are injected at rate ???.
In this experiment the liner algebra operations were run on matrixes of size ???, FFT on ??? and RK on ???.
We show results for both the routines with no error checkers as well as routines where error checkers are used with detection thresholds ???.
The data shows ???.

@@

\begin{figure}[ht!]
\centering
\includegraphics[width=1.00\columnwidth]{figs/4_1_2_Exp2_Overhead_Inputsize.png}
\includegraphics[width=1.00\columnwidth]{figs/4_1_2_Exp2_Effectiveness.png}
\includegraphics[width=1.00\columnwidth]{figs/4_1_2_Exp2_Expected_Running_Time_Needed.png}
\caption{Overhead and effectiveness of the pointer protection techniques for the ODE solver}
\label{fig:rk4_eff_ovhd}
\end{figure}

>>>>

\title{Comprehensive Algorithmic Resilience for Scientific Applications}
\author{Sui Chen, 
%\affil{Louisiana State University}
Greg Bronevetsky, Marc Casas-Guix,
%\affil{Lawrence Livermore National Laboratory}
Lu Peng
%\affil{Louisiana State University}
}

@@

\subsection{Alternating Direction Method of Multipliers (ADDR)}
\label{sec:apps:lasso}
The ADDR method solves under-unconstrained linear problems $Ax=b$ for $x$ ($A$ has fewer rows than columns) while minimizing the function $f(x)$.
This is an iterative algorithm that spends most of its time in the linear algebra operations, matrix-matrix multiplication, matrix-vector multiplication, rank-k update and Cholesky factorization, using the implementations of these routines from the GNU Scientific Library~\cite{gsl:2011}.
We focused on a specific application that uses the ADDR method to solve Lasso problems, which are defined as minimizing the function $\frac{1}{2} \left|| Ax - b \right||_2^2 + \lambda*\left|| x \right||_1$.
In our experiments we considered linear systems of size $\{40, 80, 120, ..., 4000\} \times 500$ that were generated by sampling a normal distribution (According to our experiments, the runing time of this solver is not dependent on the kind of distribution the inputs are sampled from.

@@

The DRC program generates filters for HiFi audio systems to compensate for the reflection of sounds in a room, using impulse response measurements of the audio equipment and considering the positions of the listeners. The inputs are stored in PCM format, which is an array of 32-bit floating point numbers representing the samples at each sampling time step.
Although the algorithm is divided into many phases the most computationally intensive step consists of the GSL's implementation of the Fast Fourier Transform. This program accepts user-controllable inputs to decide the width and number of iterations of FFT transforms it performs in these various phases. Different configurations would load the CPU for different lengths of time, virtually equivalent to inputs of different sizes.
\greg{What inputs?}

@@

\paragraph{Matrix-vector multiplication (MVM)}:
The MVM algorithm $Ax=b$ is checked using a similar identity $(x^TA)x = x^T(Ax) = x^Tb$.
The product $x^TA$ is a checksum of the columns of matrix $A$ and when $x$ is the vector of 1's, the value used in our experiments, it is the sum of $A$'s columns.
Although the computation of $x^TA$ involves additions rather than multiplications the complexity of computing $x^TA$ is $O(n^2)$, the same as the original MVM operation, making this check significantly less efficient.
%While this cost can be reduced significantly by computing $x^TA$ once and reusing it in subsequent multiplications by the same matrix, we did not consider it in this paper because MVM took up little application execution time.
\sui{However, for applications where matrix re-use occurs frequently, for example, a transformation matrix is used to transform many vectors, re-using the column sums would reduce the overhead of this check substantially.}

@@

\paragraph{Fast Fourier Transform (FFT)}:
The FFT algorithm decomposes a function into a sum of sine waves of different frequencies: $f(x) = \sum_{n=0}^{N-1} x_n e^{-i2\pi k n / N}$ for some constant $k$.
We check the results of FFT by using Parseval's theorem, which states that $\sum_{n=0}^{N-1} \left| x[n] \right|^2 = \frac{1}{N} \sum_{k=0}^{N-1} \left| X[k] \right|^2$, where $x$ is the original function and $X$ is its transform. GSL implements optimized radix 2, 3, 4, 5, 6 and 7 subtransform routines as well as a general radix-N FFT routine. The subtransforms are optimized and consume $O(n log(n))$ time and the general one consumes $O(n^2)$ time. Therefore, GSL has a scheduling algorithm that factorizes the length of the input FFT, running the subtransforms whenever possible, using the fallback general radix-N routine only for the remaining factors.
The theorem applies to FFT of any radix.
This means that the total energy of the original function is preserved by the transform and provides an $O(n)$ time check for the $O(n log(n))$ FFT algorithm.

\paragraph{FIR Filter Generation (FIR)}:
The FIR filter generation algorithm used in DRC generates a series of samples over the sinc function $f(x)=\frac{sin(x)}{x} (f(x)=1\ when\ x\ is\ 0)$ and modulates it with a Blackman window.
The sinc function has a nice property that $\int_{-\infty}^{\infty} sinc(x)dx = 1$. The Blackman window - modulated output of this routine
preserves this same property for most of the time, and therefore we have the checker compute the sum of the output array and compare that with 1.
Computing the sum requires merely $O(n)$ additions while this non-iterative FIR generation algorithm takes $O(n)$ trigonometric function evaluations, which much more complex than addition. Results have shown
the overhead of this checker is very small (less than 5\%).
Since this algorithm has only output and no input and therefore it does not need to have to be reinforced with input backup as other routines do.

\paragraph{Runge-Kutta PDE Solver (RK)}:
This algorithm uses the 4-th order Runge-Kutta method for solving Ordinary Differential Equations of the form $\frac{dy}{dx} = f(y, x)$.
It works by advancing the variable $x$ by steps of size $h$ and using estimates of the derivative $\frac{dy}{dx}$ at each point $x$ to find the value of $y$ at the next point $x+h$.
The RK method in GSL uses adaptive step-size control (user-selectable) where it simultaneously performs the RK method with two step sizes $h$ and $\frac{h}{2}$ (more precise).
If the difference between the results obtained with the two step sizes are larger than a threshold, the algorithm switches to the smaller step size to maintain accuracy and resumes.
Since the adaptive step control algorithm can also detect errors we utilize it to this end in our work.

@@

We now evaluate the effectiveness of our algorithmic detectors when errors are injected.
Figure~\ref{fig:algo_err_dist} shows the fraction of entries in the outputs of each routine (y-axis) that have an error of a given magnitude (x-axis) as error magnitudes span from $10^{-14}$ to $10^{-1}$ and errors are injected at three rates for each routine. With a total of 5 input sizes and three fault rates, we have a combination of 15 error rates, from 0.1 error per run to 100 errors per run for each routine.
In this experiment the liner algebra operations were run on matrixes of size 100, 150, 200, 250, 300, 500, FFT on input array sizes 512K, 1M, 2M, 4M and 8M and RK on ???.
We show results for both the routines with no error checkers as well as routines where error checkers are used with detection thresholds ???.
The data shows ???.

@@

\begin{figure}[ht!]
\centering
\includegraphics[width=1.00\columnwidth]{figs/4_1_2_Exp2_Effectiveness.png}
\includegraphics[width=1.00\columnwidth]{figs/4_1_2_Exp2_Expected_Running_Time_Needed.png}
\caption{Overhead and effectiveness of the pointer protection techniques for the ODE solver}
\label{fig:rk4_eff_ovhd}
\end{figure}
